### 2023-04-12 submission/manuscript.tex
Update Rmd file with refactored data/code

@@

Since the development of sequencing technologies such as those provided
by 454 and Illumina, microbiome researchers have struggled to produce a
consistent number of sequences from each sample in a dataset. It is
common to observe more than 10-fold variation in the number of sequences
per sample {[}XXXXX{]}. Regardless of the source of this variation,
researchers desire approaches to control for uneven sampling effort. Of
course, this desire is not unique to microbiome research and is a
challenge faced by all community ecologists. Common approaches to
controlling uneven sampling efforts have included use of proportional
abundance (i.e., relative abundance), normalization of counts, and
rarefaction.

In 2014 Paul McMurdie and Susan Holmes published their ``Wast not, want
not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
Computational Biology. This paper has had a significant impact on the
approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,300 times as of January 2023. Anecdotely, I have received
correspondence from researchers over the past 10 years asking how to
address critiques from reviewers who criticize my correspondents'
analysis for rarefying. I have also received these types of comments
from reviewers, specifically in regards to a preprint that I posted in
202X in regards to my critique of the practice of removing rare taxa
from analyses. In the process of preparing a manuscript investigating
rarefaction and other approaches to control for uneven sequencing
effort, I decided to reassess the WNWN study including their
definitions, simulations, and analyses.

Re-running the R code that the authors published as Protocol S1, I
noticed that the step that purported to rarefy the data only performed
one subsampling of the data (Lines 404 through 416 of
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
This caused me to re-inspect how McMurdie and Holmes defined
``rarefying'' in the following quoted text from their paper:

\begin{quote}
Instead, microbiome analysis workflows often begin with an ad hoc
library size normalization by random subsampling without replacement, or
so-called rarefying {[}17{]}--{[}19{]}. There is confusion in the
literature regarding terminology, and sometimes this normalization
approach is conflated with a non-parametric resampling technique ---
called rarefaction {[}20{]}, or individual-based taxon re-sampling
curves {[}21{]} --- that can be justified for coverage analysis or
species richness estimation in some settings {[}21{]}, though in other
settings it can perform worse than parametric methods {[}22{]}. Here we
emphasize the distinction between taxon re-sampling curves and
normalization by strictly adhering to the terms rarefying or rarefied
counts when referring to the normalization procedure, respecting the
original definition for rarefaction. Rarefying is most often defined by
the following steps {[}18{]}.
\end{quote}

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. This has also
  been called the rarefaction level {[}17{]}, though we will not use the
  term here.
\item
  Discard libraries (microbiome samples) that have fewer reads than
  N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\end{enumerate}
\end{quote}

\begin{quote}
Often N\textsubscript{L,m} is chosen to be equal to the size of the
smallest library that is not considered defective, and the process of
identifying defective samples comes with a risk of subjectivity and
bias. In many cases researchers have also failed to repeat the random
subsampling step (3) or record the pseudorandom number generation
seed/process --- both of which are essential for reproducibility.
\end{quote}

It is unfortunate that McMurdie and Holmes used the term rarefying here
and throughout their manuscript. The authors were correct to state that
the distinction between ``rarefying'' and ``rarefaction'' is confusing
and leads to their conflation. In my experience, subsequent researchers
have conflated the results of this study of the effects of rarefying
data with rarefaction. Adding to the confusion is that the papers cited
in their first sentence either do not use the words ``rarefy'' or
``rarefying'' or use them interchangably with ``rarefaction''. In
hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize
the distrinction between rarefying and rarefaction. However, because
they seem to have coined a new meaning for rarefying, they seem to have
only added to the confusion by using the generally used verb form of
rarefaction. Further confusion comes from the author's admonition in the
final sentence that some researhcers have failed to repeat the
subsampling step. To most scientists, repeating the subsampling step is
rarefaction. My preference is to use subsampling as the term describing
the process they refer to as rarefying. In other words rarefaction with
a single randomization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}.
\item
  Discard samples that have fewer reads than N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\item
  Compute the desired metric (e.g., richness, Shannon diversity,
  Bray-Curtis distances) using the subsampled data
\item
  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000)
\item
  Compute summary statistics using values generated from the subsampled
  data
\end{enumerate}

With this approach, rarefaction can be performed using any alpha or beta
diversity metric. Furthermore, the procedure could also be used for
hypothesis tests of differential abundance; however, thought would need
to be given to how to synthesize the results of these tests across a
large number of replications. Researchers are encouraged to report the
minimum library size as well as the number of iterations used in their
analysis. It is important to note that this procedure generates
substantially different results to those obtained without accounting for
uneven sampling effort or using relative abundances, normalized counts,
compositional data transformatins, and variance stabilization
procedures. I have explored the differences in results obtained using
the diversity of approaches for controlling for uneven sampling effort;
rarefaction, as described here, outperforms the other approaches
{[}XXXXX{]}.

Additional issues * Removed rare/patchy taxa from dataset after
subsampling the data - * Removed samples only from subsampled datasets,
which they admitted would detrimentally affect performance * Model for
defining differential abundance is weird

In this review of WNWN will reassess three things\ldots{} * Toy example
* Cluster analysis * Differential abundance

@@

\hypertarget{toy-example}{%
\subsubsection{Toy example}\label{toy-example}}

\hypertarget{cluster-analysis}{%
\subsubsection{Cluster analysis}\label{cluster-analysis}}

\hypertarget{differential-abundance}{%
\subsubsection{Differential abundance}\label{differential-abundance}}

>>>>

\begin{verbatim}
## Warning: package 'ggplot2' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'tibble' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'dplyr' was built under R version 4.2.3
\end{verbatim}

@@

Since the development of sequencing technologies such as those provided
by 454 and Illumina, microbiome researchers have struggled to produce a
consistent number of sequences from each sample in a dataset. It is
common to observe more than 10-fold variation in the number of sequences
per sample {[}XXXXX{]}. Regardless of the source of this variation,
researchers desire approaches to control for uneven sampling effort. Of
course, this desire is not unique to microbiome research and is a
challenge faced by all community ecologists. Common approaches to
controlling uneven sampling efforts have included use of proportional
abundance (i.e., relative abundance), normalization of counts, parameter
estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published their ``Wast not, want
not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
Computational Biology {[}XXXXX{]}. This paper has had a significant
impact on the approaches that microbiome researchers use to analyze 16S
rRNA gene sequence data. According to Google Scholar, this paper has
been cited more than 2,300 times as of January 2023. Anecdotely, I have
received correspondence from researchers over the past 10 years asking
how to address critiques from reviewers who criticize my correspondents'
analysis for rarefying (e.g., see
\href{https://twitter.com/inanna_nalytica/status/1264679859672006656}{this
Twitter thread}). I have also received these types of comments from
reviewers, specifically in regards to a preprint that I posted in 20XX
in regards to my critique of the practice of removing rare taxa from
analyses {[}XXXXX{]}. In the process of responding to these critiques
and preparing a manuscript investigating rarefaction and other
approaches to control for uneven sequencing effort, I decided to
reassess the WNWN study including their definitions, simulations, and
analyses.

\hypertarget{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}{%
\subsubsection{Confusion regarding what is meant by ``rarefying'' and
``rarefaction''}\label{confusion-regarding-what-is-meant-by-rarefying-and-rarefaction}}

As I attempted to reproduce the results of WNWN, I noticed that the step
that purported to rarefy the data only performed one subsampling of the
data (Lines 404 through 416 of
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
This caused me to re-inspect how McMurdie and Holmes defined
``rarefying'' in the following quoted text from their paper:

\begin{quote}
Instead, microbiome analysis workflows often begin with an ad hoc
library size normalization by random subsampling without replacement, or
so-called rarefying {[}17{]}--{[}19{]}. There is confusion in the
literature regarding terminology, and sometimes this normalization
approach is conflated with a non-parametric resampling technique ---
called rarefaction {[}20{]}, or individual-based taxon re-sampling
curves {[}21{]} --- that can be justified for coverage analysis or
species richness estimation in some settings {[}21{]}, though in other
settings it can perform worse than parametric methods {[}22{]}. Here we
emphasize the distinction between taxon re-sampling curves and
normalization by strictly adhering to the terms rarefying or rarefied
counts when referring to the normalization procedure, respecting the
original definition for rarefaction. Rarefying is most often defined by
the following steps {[}18{]}.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. This has also
  been called the rarefaction level {[}17{]}, though we will not use the
  term here.
\item
  Discard libraries (microbiome samples) that have fewer reads than
  N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\end{enumerate}

Often N\textsubscript{L,m} is chosen to be equal to the size of the
smallest library that is not considered defective, and the process of
identifying defective samples comes with a risk of subjectivity and
bias. In many cases researchers have also failed to repeat the random
subsampling step (3) or record the pseudorandom number generation
seed/process --- both of which are essential for reproducibility.
\end{quote}

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
here and throughout their manuscript. The authors were correct to state
that the distinction between ``rarefying'' and ``rarefaction'' is
confusing and leads to their conflation. In my experience, subsequent
researchers have conflated the results of this study of the effects of
rarefying data with rarefaction of data. As an example, Willis (XXXXX)
describes problems with rarefaction rather than rarefying data when
citing WNWN in her paper proposing alternatives to rarefaction for use
with alpha diversity data:

\begin{quote}
Unfortunately, rarefaction is neither justifiable nor necessary, a view
framed statistically by McMurdie and Holmes (2014) in the context of
comparison of relative abundances.
\end{quote}

Adding to the confusion is that the papers cited in the first sentence
of the quote I WNWN included above either do not use the words
``rarefy'' or ``rarefying'' or use them interchangably with
``rarefaction''. In hindsight, as shown in the quoted text, McMurdie and
Holmes do emphasize the distinction between rarefying and rarefaction.
However, because they seem to have coined a new meaning for rarefying,
they seem to have only added to the confusion by using the generally
used verb form of rarefaction. Further confusion comes from the author's
admonition in the final sentence that some researchers have failed to
repeat the subsampling step. To most scientists, repeating the
subsampling step is rarefaction. My preference is to use subsampling as
the term describing the process they refer to as rarefying. In other
words rarefaction with a single randomization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. Researchers are
  encouraged to report the value of N\textsubscript{L,m}.
\item
  Discard samples that have fewer reads than N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\item
  Compute the desired metric (e.g., richness, Shannon diversity,
  Bray-Curtis distances) using the subsampled data
\item
  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000).
  Researchers are encouraged to report the number of iterations.
\item
  Compute summary statistics using values generated from the subsampled
  data
\end{enumerate}

This definition aligns well with how rarefaction was originally defined
for comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths. It is
important to note that this procedure generates substantially different
results to those obtained without accounting for uneven sampling effort
or using relative abundances, normalized counts, compositional data
transformatins, variance stabilization procedures, or estimation
techniques. I have explored the differences in results obtained using
the diversity of approaches for controlling for uneven sampling effort;
rarefaction, as described here, outperforms the other approaches
{[}XXXXX{]}.

With this more general approach to rarefaction, rarefaction can be
performed using any alpha or beta diversity metric. This strategy has
been widely used by my research group and others. The procedure outlined
above could also be used for hypothesis tests of differential abundance;
however, thought would need to be given to how to synthesize the results
of these tests across a large number of replications.

\hypertarget{description-of-simulation-a-from-wnwn}{%
\subsubsection{Description of ``Simulation A'' from
WNWN}\label{description-of-simulation-a-from-wnwn}}

McMurdie and Holmes analyzed the effect of rarefying and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign samples to one of two
clusters using simulated data used to generate 40 samples from each of
two distributions. A variety of approaches were used to calculate
distances between the samples those distances were used as input to
partitioning around mediods (PAM) clustering with two groups. The
accuracy of the cluster assignment was used as the metric to assess
performance. This analysis was performed in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in the
original paper. Line numbers from this file will be refernced with an
``L'' as a prefix.

The two distributions were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
generate a fecal and ocean template distribution, the authors included
any operational taxonomic unit (OTU) that appeared in more than one of
the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by
how many of the 7 samples the OTUs were observed in followed by their
total abundance across all 7 samples (L139). From this sorted list they
identified the identifiers of the first 2000 OTUs (L66). Returning to
the 7 samples they selected the 2000 most common and abundant OTUs and
pooled the abundances of the fecal and ocean samples separately to
create two templates (L144, L159-160, L197-198).

Next, the fecal and ocean templates were mixed in 8 different fractions
to generate two community types that differend by varying effect sizes
(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220). To simulate
the variation in sequencing depth across the 80 samples, they normalized
he number of sequences from each of the \texttt{gp\_n\_samples} samples
in the GlobalPatterns dataset so that the median number of sequences
(N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000, 5,000, or
10,000 sequences (L324-325). They then randomly sampled the
\texttt{gp\_n\_samples} normalized sequencing depths to generate 80
sampling depths. From each community type, they simulated 40 samples by
sampling to the desired number of reads (L73, L230-233 and L326-327).
Each simulation condition was repeated 5 times (L85). This resulted in
160 simulations (8 effect sizes x 4 median sampling depths x 5
replicates = 160 simulations). Finally, they removed rare and low
prevalence OTUs in two steps. First, they removed any OTUs whose total
abundance was less than 3 across all 80 samples and that did not appear
in at least 3 samples (L368-386). Second, they removed any OTUs that did
not have more than 1 sequence in more than 5\% of the 80 samples (i.e.,
4 samples) and that did not have a total abundance across the 80 samples
greater than one half of the number of samples in each community type
(i.e., 20) (L523-538, L551).

\hypertarget{critique-of-the-original-simulation-design}{%
\subsubsection{Critique of the original simulation
design}\label{critique-of-the-original-simulation-design}}

Although all simulations represent an artificial representation of
reality and can be critiqued, eleven elements of the design of
Simulation A warrant further review.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, which caused
  results to be sensitive to random number generator seed
\item
  The average sizes of the libraries were small by modern standards
\item
  DeSeq-based variance stabilization was used with inappropriate
  distance calculation methods
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was poorly chosen
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which could distort the shape of the communities
\item
  No test of effect of transformation methods to account for effects of
  uneven sample size between treatment groups
\item
  Clustering accuracy was used rather than direct comparisons beta
  diversity
\item
  No consideration of effects of transformations on alpha diversity
  metrics
\end{enumerate}

These points will serve as an outline for the Results section. After
replicating the original simulations, these points will be evaluated to
reassess whether subsampling or rarefaction are ``inadmissible''.

@@

\hypertarget{replication-of-wnwn-simulations-and-results}{%
\subsubsection{Replication of WNWN simulations and
results}\label{replication-of-wnwn-simulations-and-results}}

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of the original paper
using the authors original code and my own. I created a Conda
environment that used the R version and package versions that were as
close as possible to those used in the original paper. Because of the
slight differences in packages, it was necessary to apply several
patches to the original R-flavored markdown file to get the document to
render. I was able to generate a figure similar to that presented as
Figures 4 and 5 of the original paper. My results are shown in Figures
S2 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf})
and S3
(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf}) of
this paper, respectively. The differences in results are likely due to
differences in software versions and operating systems. It is also worth
noting that their versions of the two figures differ from those included
in Protocol S1 within the rendered html file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file. Regardless of the
differences, the results are qualitatively similar.

\hypertarget{simulated-conditions-were-only-replicated-5-times-each}{%
\subsubsection{1. Simulated conditions were only replicated 5 times
each}\label{simulated-conditions-were-only-replicated-5-times-each}}

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DEeqVS transformed data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies are unlikely to be symmetric around a
mean at larger accuracy values the median and 95\% confidence intervals
or intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment as above. This was necessary
since the number of simulated conditions because of this change
increased 20-fold. Such intense data processing was not practical within
a single R-flavored markdown document. Again, the observed results were
qualitiatively similar to those generated using the single R-flavored
markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards}{%
\subsubsection{2. The average sizes of the libraries were small by
modern
standards}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards}}

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
XXXXXXXX{]}). Therefore, I included an additional average depth of
sampling value of 50,000 sequences with the four average sequencing
sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000). Additional
sequencing coverage would be expected to result in more robust distance
values since there would be more information represented in the data.
Indeed, the added sampling depth showed higher accuracy values at lower
effect sizes for the combinations of normalization methods and distance
calculations (Figure S4 (\textbf{pam\_subsample15\_fig\_4.pdf})).
Increased sequencing coverage also resulted in improved clustering
accuracy for lower effect sizes when the library size minimum quantile
was decreased (Figure S5 (\textbf{pam\_subsample\_fig\_5.pdf})). I will
revisit the choice of the library size minimum quantile below.

\hypertarget{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}{%
\subsubsection{3. DeSeq-based variance stabilization was used with
inappropriate distance calculation
methods}\label{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}}

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned 0.51 and 1.00. The
Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DeSeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated a warning message when passing the DeSeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the weighted UniFrac calculator function did not
generate a similar warning message, negative values would also result in
similarly meaningless distances. Both are due to the fact that the
calculators sum the counts of each OTU in both samples being compared.
In contrast, a Euclidean distance does not use a similar sum, but sums
the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I
quantified the fraction of negative values in the OTU matrix from each
simulation and counted the number of simulations where the transformed
OTU table had at least one negative value (Figure
\textbf{deseq\_negative\_value.pdf}). In general the fraction of
negative OTU counts increased with effect size, but decreased with
sequencing effort. The fraction of simulations with at least onee
negative value increased with effect size and sequencing effort. The
high frequency of negative OTU counts resulted in highly variable
Bray-Curtis and weighted UniFrac values. It is likely that because the
WNWN analysis only used 5 replicates that the large variation in
accuracies at high effect sizes was missed initially. For the rest of
this reanalysis study, I will only report results using the DeSeq-based
variance stabilization transformation with the Euclidean distance.

\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}{%
\subsubsection{4. A single subsampling of each dataset was evaluated
rather than using
rarefaction}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}}

As noted above, the original jargon that was used in WNWN was confusing
to many who conflated rarefying with rarefaction. One problem with a
single subsampling of a community is that it is unlikely to obtain a
representative sampling of each community. A more robust analysis would
have used rarefaction rather than a single subsampling of the data since
it would have averaged across random subsamplings, which individually
would be unlikely to represent the overall composition of the
communities. Rather than being guilty of ``omission of available valid
data'', rarefaction uses all of the available data. To compare
subsampling and rarefaction, I removed the 15\% of samples with the
lowest number of sequences for each of the 100 simulated datasets and
compared the distances from a single subsampling to rarefaction using
the distance calculations shown in the original Figure 4. This analysis
revealed two benefits of rarefaction. First, the median distances
generated by rarefaction was always as large or larger than those from a
single subsample (Figure \textbf{subsample\_rarefaction\_compare.pdf}).
In general, the difference was most pronounced for smaller average
library sizes and at smaller effect sizes. The unweighted UniFrac
distances were most impacted by the use of rarefaction over subsampling.
Second, the intraquartile ranges for the distances generated by
rarefaction were generally smaller than those by subsampling and showed
similar trends to the difference in the median distances (Figure
\textbf{subsample\_rarefaction\_compare.pdf}). The intraquartile ranges
for Bray-Curtis, Euclidean, and unweighted UniFrac distances were
actually larger by rarefaction than by subsampling at small effect sizes
and average library sizes; however at larger values the intraquartile
range by subsampling was larger than by rarefaction for these distance
calculations. Because rarefaction incorporates more of the data and
generally performed better than subsampling, the remainder of this
analysis will report results using rarefaction rather than by
subsampling except when noted.

\hypertarget{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}{%
\subsubsection{5. Results using PAM clustering were not directly
compared to those of K-means and hierarchical
clustering}\label{results-using-pam-clustering-were-not-directly-compared-to-those-of-k-means-and-hierarchical-clustering}}

The clustering accuracy measurements in the body of the manuscript were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchichal clustering. Although the data were not
displayed in a manner that lent itself to direct comparison, close
inspection of the rendered figures suggests that PAM may not have been
the optimal choice in all situations. Rather, K-means clustering may
have been prefered. Because the accuracies were the smallest at lower
effect sizes, I focused my comparison at the effect size of 1.15. For
each set of 100 replicated simulated datasets, I compared the clustering
accuracy across clustering methods to see how often each clustering
method resulted in the highest accuracy (Figure
\textbf{compare\_cluster\_methods.pdf}). Indeed, K-means clustering
performed better than the other methods. Among all combinations of
normalization methods, distance calculations, and read depths, PAM
clustering resulted in clustering accuracies as good or better than the
other methods in 49.92\% of the randomizations (Figure
\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations. HClust
was at least as good as the other methods for 44.32\% of the
randomizations. I specifically compared the clustering accuracies using
rarefaction for each of the distance calculations methods using PAM and
K-means clustering. Among the 30 combinations of distance calculations
and read depths, K-means performed better than PAM in 29 cases with PAM
doing better in the 1 other case when calculating distances with
Euclidean using 10,000 sequences. When using subsampled data, K-means
clustering performed better than PAM in each case. Because K-means
clustering did so much better than PAM clustering in the simulated
conditions, I will use K-means clustering for the remainder of this
study.

\hypertarget{subsampling-removed-the-smallest-15-of-the-samples}{%
\subsubsection{6. Subsampling removed the smallest 15\% of the
samples}\label{subsampling-removed-the-smallest-15-of-the-samples}}

In WNWN, the authors quantified the tradeoff between sampling effort,
the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
sampling effort varied by distance metric, transformation method, and
sampling effort, they removed samples whose number of sequences was less
than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\% (Figure S2). Because the true result would not known \emph{a
priori} in microbiome studies, it would be impossible for researchers to
conduct a sensitivity analysis comparing the tradeoffs between
sequencing depth, sample number, and clustering accuracy to select a
sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
counts requires an arbitrary selection of a library size minimum that
affects downstream inference'' (page 8, column 1, point 3), in acutal
microbiome studies the selection of a sampling depth is not as arbitrary
as the authors claim. Rather, to avoid p-hacking, researchers pick a set
of criteria where they will include or exclude samples prior to testing
their data.}{]} The differences in clustering accuracy between
subsampling and rarefaction and using PAM and K-means clustering
indicated that it was necessary to reassess the tradeoff between the
library size minimum quantile and clustering accuracy. When using
rarefaction, K-means clustering, and only considering conditions with
2,000 or more sequences, there was not a condition where setting a
higher threshold resulted in a better accuracy than using all of the
samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
that for modern sequencing depths, using the full datasets with
rarefaction and K-means clustering resulted in accuracies that were
typically better than those observed when removing the smallest 15\% of
the samples from each simulated dataset. When the original Figure 4 was
recast with these approaches, Rarefaction performed at least as well as
any of the other tranformations with each distance calculation, except
for with the Poisson distance (Figure
\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
larger effect sizes, K-means clustering did not perform as well for some
combinations of normalization methods and distance calculations (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

\hypertarget{the-distribution-of-library-sizes-was-poorly-chosen}{%
\subsubsection{7. The distribution of library sizes was poorly
chosen}\label{the-distribution-of-library-sizes-was-poorly-chosen}}

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
median of 1,106,849.00 sequences per dataset. The datasets ranged in
sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
40.16-fold difference. Rather than representing a typically observed
distribution of sequencing depths that would be skewed right, the
sampling distribution was normally distributed (Shapiro-Wilk test of
normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
From these simulations it is unclear how sensitive the various
transformations and distance calculations are to a skewed distribution.
A second limitation of this sampling distribution is that it only
contained 26 unique sampling depths such that each sampling depth would
have been re-used an average of 3.08 times in each simulation. Yet, it
is unlikely for a real sequence collection to have duplicate sequencing
depths.

I created a new set of simulations that would focus on solely on the
effect of the shape of the distribution on the results to reassess the
WNWN results in the context of a more typical distribution of sample
sizes. I created a simple sequencing depth distribution where there were
80 depths logarithmically distributed between the minimum and maximum
sequencing depths of the GlobalPatterns dataset (Figure
\textbf{\emph{distribution\_shape.pdf}}). The median of this
distribution was 372,040.00 and the mean was 629,825.00. When I
regenerated the original Figures 4 and 5 using the log-distributed
sequencing effort distribution, the differences in transformation
methods were more apparent (Figure
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
distance calculators, rarefying the data to the size of the smallest
dataset yielded accuracies that were at least as good as the other
methods across effect sizes and median sequencing depths. The difference
was most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
observed using the log-distributed sample sizes was at least as good as
those obtained using the GlobalPatterns-based distribution
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of
number of samples and the distribution of their sequencing depths in the
context of controlling for uneven sampling effort is explored in far
greater detail in an another analysis using sequencing depths observed
in biological samples {[}XXXXXX{]}.

\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}{%
\subsubsection{8. A filtering step was applied to remove rare taxa from
the simulated datasets, which could distort the shape of the
communities}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets-which-could-distort-the-shape-of-the-communities}}

McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
count data is statistically inadmissible} because it requires the
omission of available valid data'' (emphasis in original). Thus it is
strange that they argue against removing data when
rarefying/subsampling, but accept removing rare and low-prevalence OTUs
prior to normalizing and analyzing their simulated communities. This
practice has become common in microbiome studies and is the standard
approach in tools such as dada2, unoise, and deblur {[}XXXXXXXX{]}.
However, my previous work has shown that rare sequences from a poorly
sequenced sample often appear in more deeply sequened samples suggesting
that they are not necessarily artifacts. Furthermore, removing rare
sequences alters the structure of communities and has undesirable
effects on downstream analyses {[}XXXXXXXX{]}.

Although my previous work does an extensive analysis of the effects of
removing rare sequences, I wanted to explore the effect of filtering in
the context of the WNWN simulation framework. For each of the filtered
and non-filtered OTU tables I calcualted the absolute value of the
difference in accuracy between each distance calculation following the
normalization procedure (Figure \textbf{compare\_filter\_accuracy.pdf}).
With the exception of the Weighted UniFrac distances, each of the
distance calculations and normalization procedures were sensitive to the
filtering. The rarefaction data tended to be sensitive to filtering at
small effect sizes. Distances generated using raw counts, DeSeq variance
stabilization, and Upper Quartile Log Fold Change tended to be more
sensitive to filtering at larger effect sizes. When using relative
abundance data, Bray-Curtis distances were sensitive to filtering at
small effect sizes and Unweighted UniFrac distances were sensitive at
large effect sizes. These trends appear to be driven by the dependence
of the distance calculation on low abundance taxa. More surprising than
the effect of filtering on the mean absolute difference in clustering
accuracy was the wide variation in accuracies at each effect size. Among
the different normalization methods, the accuracies calculated using
rarefaction had the narrowest 95\% confidence interval for all distance
calculations except for calculating Unweighted UniFrac distances. For
these distances, relative abundance had the narrowest range at small
effect sizes; at larger effect sizes, rarefaction had the narrowest
range. Again, these trends appear to be driven by the dependence on low
frequency taxa in Unweighted UniFrac, which is dependent on the presence
or absence of taxa rather than their abundance. Given my previous work
and the large variation caused by removing rare taxa, OTU filtering
should not be performed in microbiome analyses.

\hypertarget{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}{%
\subsubsection{9. No test of effect of transformation methods to account
for effects of uneven sample size between treatment
groups}\label{no-test-of-effect-of-transformation-methods-to-account-for-effects-of-uneven-sample-size-between-treatment-groups}}
